Communicating with convexity

### 2017-11-03 article_cvxpy.tex
Corrections reccomended by referee

\subsection{Convex functions}
%If a set $C$ is \textit{convex}, then for any two points $x,y\in C$, any point $z$ along the $x,y$ line must also $\in C$ \cite[p.23]{cvxpybook}. More formally, for $x,y\in C$ and $\theta\in [0,1]$:
%\begin{align}
%\theta x + (1-\theta)y\in C.
%\end{align}

\noindent A function $f$, where $f:\mathbb{R}^n \rightarrow \mathbb{R}$, is \textit{convex} if $\forall x,y\in \textbf{dom} f$ and $0 \leq \theta \leq 1$ \cite[p.67]{cvxpybook}:
\begin{align}
f(\theta x + (1-\theta)y) \leq \theta f(x)+(1-\theta)f(y).
\end{align}
This means that the function is less than or equal to linear, as shown in Fig.~\ref{fig:convex}. %A concave function can be made convex by the operation $f\to-f$.

@@

\begin{figure}[h!]
\includegraphics[width=0.9\linewidth]{convex_function.pdf}
\caption{\label{fig:convex}A chord showing convexity of the function $x^4$.} 
\end{figure}

\subsection{Disciplined convex programming}
\noindent In order to solve convex optimisation problems, we used \texttt{cvxpy}, a symbolic programming module for \textit{Python}\cite{cvxpy}. It uses a set of rules, called \textit{disciplined convex programming} (DCP), to determine whether a function is convex. This is implemented using predefined classes containing functions with their curvature and sign, and using general composition theorems from convex analysis \cite{dcp}. If DCP rules can be applied then interior point methods guarantee an optimal solution. Interior point methods work by applying Newtons algorithm to sequences of equally constrained problems. The algorithm works by picking a step and direction to travel in the interior of the feasible convex set, such that the solution progresses towards optimality, for more details on barrier methods and primal-dual methods see \citep[p.561]{cvxpybook}.
\section{Example: power minimisation in communications}
\noindent To demonstrate the applicability of \texttt{cvxpy} to  a real-world example, we consider a system of $n$ transmitters each of power $p_i$ and $m$ receivers, all in $2$D Euclidean space\cite{shannon1949}. In this problem we are constrained to having a minimum signal-to-interference-plus-noise ratio (SINR) at each receiver; the strength of the desired signal, $S$, relative to the interference power, $I$, plus the background noise, $\sigma$, at a receiver. How can we minimise the total power consumption, $P$, of transmitters, yet achieve this minimum SINR, $\gamma_0$, for all receivers? This question is relevant to telecoms companies who want to offer a service at a minimum quality standard. We formulate the problem with a given square path gain matrix, $G$, background noise level $\sigma$ and a maximum power constraint $P_{\text{max}}$ of each transmitter:
\begin{align*}
&\underset{p}{\text{minimise}} \quad &&\sum_j p_j\\
&\text{subject to} \quad &&p_j \leq P_{\max}\\
& \quad &&p_j \geq 0\\
& \quad &&\gamma_i \geq \gamma_0.&&
\end{align*}

The desired signal for receiver $i$ is $S_i = G_{ik}p_k$, while the interference at $i$ is $I_i = \sum_{j\neq k}G_{ij}p_j$. However DCP does not allow division of the optimisation variable $p_i$ so  
\begin{equation*}
  \gamma_i \geq \gamma_0\Longleftrightarrow  \frac{S_i}{\sigma_i + I_i}\geq \gamma_0, \quad \forall i
\end{equation*}
was rearranged to,
\begin{equation*}
S_i-\gamma_0(\sigma_i + I_i)\geq 0, \quad \forall i
\end{equation*} which is affine, and hence a DCP function.

\subsection{Path gain}
\noindent The path gain $G_{ij}$ represents the proportion of power that reaches receiver $i$ from transmitter $j$. Supposing that the desired signal to receiver $i$ is from transmitter $k$, the power received is,
\begin{equation}
p_{ik}^{\text{rec}} = G_{ik}p_k.
\end{equation}
Denoting the SINR at $i$ by $\gamma_i$, we have
\begin{equation}
\gamma_i = \frac{G_{ik}p_k}{\sigma_i+\sum_{j\neq k}G_{ij}p_j}
=\frac{S_i}{\sigma_i+I_i}.
\end{equation}
In a physical context, the path gain from transmitter $j$ to receiver $i$, as shown in Fig.~\ref{fig:pathgain}, will depend on the distance, $d_{ij}$ between them. Assuming isotropic propagation from each transmitter, the fraction of power from $j$ that reaches $i$ is:
\begin{equation}
G_{ij} = k_i/d_{ij}^\alpha
\end{equation}
where $\alpha$ is the pathloss coefficient and $k$ is a proportionality constant specific to the receiver $i$.  For free space $\alpha = 2$, while in urban environments $\alpha \sim 3.5$ \cite{hata1980}. Further physical complexities, such as the stochastic effects from Rayleigh fading can be easily incorporated.

@@

\section{Outlook}
\noindent The authors hope that this article shows the utility of \texttt{cvxpy} for simple `toy' problems of real-world relevance. We believe it is an invaluable resource for researchers wanting to test the applicability of convex optimisation in new fields.

The module could also be easily incorporated into a course on optimisation, providing students with an accessible environment to practise techniques for convex optimisation. Full code for other communication examples can be found at https://github.com/cvxgrp/cvxpy/\\tree/master/examples/communications.

>>>>

\subsection{Convex functions}
If a set $C$ is \textit{convex}, then for any two points $x,y\in C$, any point $z$ along the $x,y$ line must also $\in C$ \cite[p.23]{cvxpybook}. More formally, for $x,y\in C$ and $\theta\in [0,1]$:
\begin{align}
\theta x + (1-\theta)y\in C.
\end{align}

\noindent A function $f$, where $f:\mathbb{R}^n \rightarrow \mathbb{R}$, is \textit{convex} if $\forall x,y\in \textbf{dom} f$ and $0 \leq \theta \leq 1$ \cite[p.67]{cvxpybook}:
\begin{align}
f(\theta x + (1-\theta)y) \leq \theta f(x)+(1-\theta)f(y).
\end{align}
This means that the function is less than or equal to linear, as shown in Fig.~\ref{fig:convex}. A concave function can be made convex by the operation $f\to-f$.

@@

\begin{figure}[h!]
\includegraphics[width=0.9\linewidth]{convex_function.pdf}
\caption{\label{fig:convex}The function $f(x)=x^4$ is plotted in blue. The convexity of this function can be seen by picking any two values of $x$ and noting that $f(x)$ will always lie below or on the chord connecting these two points.} 
\end{figure}

\subsection{Disciplined convex programming}
\noindent In order to solve convex optimisation problems, we used \texttt{cvxpy}, a symbolic programming module for \textit{Python}\cite{cvxpy}. It uses a set of rules, called \textit{disciplined convex programming} (DCP), to determine whether a function is convex. This is implemented using predefined classes containing functions with their curvature and sign, and using general composition theorems from convex analysis \cite{dcp}. If DCP rules can be applied then interior point methods guarantee an optimal solution. Interior point methods work by applying Newtons algorithm to sequences of equally constrained problems. The algorithm works by picking a step and direction to travel in the interior of the feasible convex set, such that the solution progresses towards optimality, for more details on barrier methods and primal-dual methods see \citep[p.561]{cvxpybook}.

\section{Example: power minimisation in communications}
\noindent To demonstrate the applicability of \texttt{cvxpy} to  a real-world example, we consider a system of $n$ transmitters each of power $p_i$ and $m$ receivers, all in $2$D Euclidean space\cite{shannon1949}. In this problem we are constrained to having a minimum signal-to-interference-plus-noise ratio (SINR or $\gamma$) at each receiver
\begin{equation}
\gamma =\frac{S}{I+\sigma}
\end{equation} 
where $S$ is the power of the desired signal, $I$ is the interference power and $\sigma$ is the background noise at a receiver.

How can we minimise the total power consumption, $P$, of transmitters, yet achieve this minimum SINR, $\gamma_0$, for all receivers? This question is relevant to telecoms companies who want to offer a service at a minimum quality standard. We formulate the problem with a given square path gain matrix, $G$, background noise level $\sigma$, a maximum power constraint $P_{\text{max}}$ of each transmitter and the fact that all transmitters must have positive power:
\begin{align*}
&\underset{p}{\text{minimise}} \quad &&\sum_j p_j\\
&\text{subject to} \quad &&p_j \leq P_{\max}\\
& \quad &&p_j \geq 0\\
& \quad &&\gamma_i \geq \gamma_0.&&
\end{align*}

The desired signal for receiver $i$ is $S_i = G_{ik}p_k$, while the interference at $i$ is $I_i = \sum_{j\neq k}G_{ij}p_j$. However DCP does not allow division of the optimisation variable $p_i$, as it cannot guarantee curvature, so  
\begin{equation*}
  \gamma_i \geq \gamma_0\Longleftrightarrow  \frac{S_i}{\sigma_i + I_i}\geq \gamma_0, \quad \forall i
\end{equation*}
is rearranged to,
\begin{equation*}
S_i-\gamma_0(\sigma_i + I_i)\geq 0, \quad \forall i
\end{equation*} which is affine, and hence a DCP function.

\subsection{Path gain}
\noindent The path gain $G_{ij}$ represents the proportion of power that reaches receiver $i$ from transmitter $j$. Supposing that the desired signal to receiver $i$ is from transmitter $k$, the power received is,
\begin{equation}
p_{ik}^{\text{rec}} = G_{ik}p_k.
\end{equation}
Denoting the SINR at $i$ by $\gamma_i$, we have
\begin{equation}
\gamma_i = \frac{G_{ik}p_k}{\sigma_i+\sum_{j\neq k}G_{ij}p_j}
=\frac{S_i}{\sigma_i+I_i}.
\end{equation}
In a physical context, the path gain from transmitter $j$ to receiver $i$, as shown in Fig.~\ref{fig:pathgain}, will depend on the distance, $d_{ij}$ between them. Assuming isotropic propagation from each transmitter, the fraction of power from $j$ that reaches $i$ is:
\begin{equation}
G_{ij} = k_i/d_{ij}^\alpha
\end{equation}
where $\alpha$ is the pathloss coefficient and $k$ is a proportionality constant specific to the receiver $i$.  For free space $\alpha = 2$, while in urban environments $\alpha \sim 3.5$ \cite{hata1980}. Further physical complexities, such as the stochastic effects from Rayleigh fading can be easily incorporated, whilst remaining within the DCP framework.

@@

\section{Outlook}
\noindent The authors hope that this article shows the utility of \texttt{cvxpy} for simple `toy' problems of real-world relevance. $\texttt{cvxpy}$ can handle problems with up to $\sim 10^3$ transmitters, however this takes $\sim 10^{4}-10^{5}\text{ms}$ to run, hence for large real-world problems a faster implementation would likely be required (e.g. C++ based library). However there is obviously a trade off in terms of ease of understanding with these packages.

We believe that $\texttt{cvxpy}$ is an invaluable resource for researchers wanting to test the applicability of convex optimisation in new fields. The module could also be easily incorporated into a course on optimisation, providing students with an accessible environment to practise techniques for convex optimisation. Full code for other communication examples can be found at https://github.com/cvxgrp/cvxpy/\\tree/master/examples/communications.
